What is an Ancestral Recombination Graph?

### 2022-03-03 paper.tex
Clarified single chr scope and unified Fig1 with text

In this paper we [turn into proper round up paragraph later]
\begin{enumerate}
\item Discuss the classical Ancestral Recombination Graph formulation,
and show how it derives from (and is limited by) the coalescent with
recombination.
\item Suggest a definition of ARGs that is free of these limitations,
where the formulation is derived from a (normally diploid) pedigree rather than from
coalescent approximations, and where the genetic
material which is transmitted is described by annotations attached to
edges rather than nodes.
\item Suggest a classification of ARG nodes that helps us to understand
the properties of the structures inferred by different methods, and illustrate
with some examples.
% Perhaps we should also mention something about simplification and unknowable
% nodes, e.g. "show how some of the nodes are not required for useful interpretation
% of ancestry".
\end{enumerate}

@@

	\draw (0, 0) -- (0, 1.5) -- (-1,1.5) -- (1,1.5);
	\draw (-1.4, 1.3) -- (-1.1, 1.3) -- (-1.1, 1.2) -- (-1.4, 1.2) -- (-1.4, 1.3);
	\draw (-1.1, 1.25) -- (-0.6, 1.25);
	\draw (0.6, 1.25) -- (0.9, 1.25);
	\draw (0.9, 1.3) -- (1.4, 1.3) -- (1.4, 1.2) -- (0.9, 1.2) -- (0.9, 1.3);
	\draw (1.2, 1.4) -- (1.2, 1.1);
	\node [label=below:{\small $x$}] at (1.2, 1.3) {};

@@

	\draw (1, 1.5) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
	\draw (-0.4, 2.85) -- (0.2, 2.85);
	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
	\draw (1.6, 2.85) -- (1.9, 2.85);
	\draw (1.9, 2.9) -- (2.2, 2.9) -- (2.2, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
	\draw (2.2, 2.85) -- (2.4, 2.85);

	\draw (-1, 1.5) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
	\node [scale=0.2,label=above left:{\small A}] at (-0.5,4.5) {A};
	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
	\draw (-0.6, 4.25) -- (-0.3, 4.25);
	\draw (-0.3, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.3, 4.2) -- (-0.3, 4.3);
	\draw (-0.5, 4.4) -- (-0.5, 4.1);
	\node [label=below:{\small $y$}] at (-0.5, 4.3) {};

@@

	% Edge annotations above each event
	\node [label=left:{\small$\{[0,L)\};1$}] at (0.2,0.6) {};
	\node [label=left:{\small$\{[0,L)\};1$}] at (4.2,0.4) {};
	\node [label=right:{\small$\{[0,L)\};1$}] at (5.8,0.4) {};
	\node [label=left:{\small$\{[0,L)\};2$}] at (5.2,1.6) {};
	\node [label=left:{\small$\{[0,v)\};1$}] at (-0.8,2.3) {};
	\node [label=right:{\small$\{[v,L)\};1$}] at (0.8,2.1) {};
	\node [label=right:{\small$\{[0,w)\};2$}] at (3.8,3.9) {};
	\node [label=left:{\small$\{[w,L)\};2$}] at (6.2,5) {};
	\node [label=right:{\small$\{[x,L)\};1$}] at (-0.2,3.8) {};
	\node [label=right:{\small$\{[v,x)\};1$}] at (1.8,4.1) {};
	\node [label=left:{\small$\{[0,v), [x, L)\};1$}] at (-0.3,5.2) {};
	\node [label=right:{\small$\{[0,v)\};2$}] at (2.8,6.4) {};
	\node [label=left:{\small$\{[x,L)\};1$}] at (-1.3,7) {};
	\node [label=right:{\small$\{[0,v)\};1$}] at (0.3,6.4) {};

@@

	% Numbers of extant ancestors and links, from top to bottom
	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
	\node[label=right:{$\binom{2}{2}$ \; $(2 L - x - w) \rho$}] at (7.5, 7.6) {};
	\node[label=right:{$\binom{4}{2}$ \; $(2 L + 2 v - x - w) \rho$}] at (7.5, 6.6) {};
	\node[label=right:{$\binom{3}{2}$ \; $(2 L + v - w) \rho$}] at (7.5, 5.7) {};
	\node[label=right:{$\binom{4}{2}$ \; $(2 L + x - v) \rho$}] at (7.5, 4.95) {};
	\node[label=right:{$\binom{5}{2}$ \; $2 L \rho$}] at (7.5, 3.8) {};
	\node[label=right:{$\binom{4}{2}$ \; $2 L \rho$}] at (7.5, 2.75) {};
	\node[label=right:{$\binom{3}{2}$ \; $2 L \rho$}] at (7.5, 1.95) {};
	\node[label=right:{$\binom{2}{2}$ \; $2 L \rho$}] at (7.5, 1.35) {};
	\node[label=right:{$\binom{3}{2}$ \; $3 L \rho$}] at (7.5, 0.6) {};

	% Gray dashed lines to visually separate holding times
	\draw[color=gray, dashed] (7.5, 1.2) -- (11.1, 1.2);
	\draw[color=gray, dashed] (7.5, 1.5) -- (11.1, 1.5);
	\draw[color=gray, dashed] (7.5, 2.4) -- (11.1, 2.4);
	\draw[color=gray, dashed] (7.5, 3.1) -- (11.1, 3.1);
	\draw[color=gray, dashed] (7.5, 4.5) -- (11.1, 4.5);
	\draw[color=gray, dashed] (7.5, 5.4) -- (11.1, 5.4);
	\draw[color=gray, dashed] (7.5, 6) -- (11.1, 6);
	\draw[color=gray, dashed] (7.5, 7.2) -- (11.1, 7.2);
\end{tikzpicture}
&
\begin{tikzpicture}
	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
	\draw (-0.1, -0.1) -- (-0.1, -0.4);
	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);

@@

	\draw (1, 1.5) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
	\draw (-0.4, 2.85) -- (0.2, 2.85);
	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
	\draw (1.6, 2.85) -- (1.9, 2.85);
	\draw (1.9, 2.9) -- (2.2, 2.9) -- (2.2, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
	\draw (2.2, 2.85) -- (2.4, 2.85);

	\draw (-1, 1.5) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
	\draw (-0.6, 4.25) -- (-0.3, 4.25);
	\draw (-0.3, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.3, 4.2) -- (-0.3, 4.3);
	\draw (-0.5, 4.4) -- (-0.5, 4.1);
	\node [label=below:{\small $y$}] at (-0.5, 4.3) {};

@@

	% Numbers of extant ancestors and links, from top to bottom
	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
	\node[label=right:{$\binom{3}{2}$ \; $3 L \rho$}] at (7.5, 7.6) {};
	\node[label=right:{$\binom{4}{2}$ \; $4 L \rho$}] at (7.5, 6.9) {};
	\node[label=right:{$\binom{5}{2}$ \; $5 L \rho$}] at (7.5, 6.3) {};
	\node[label=right:{$\binom{4}{2}$ \; $4 L \rho$}] at (7.5, 5.7) {};
	\node[label=right:{$\binom{5}{2}$ \; $5 L \rho$}] at (7.5, 4.95) {};
	\node[label=right:{$\binom{6}{2}$ \; $6 L \rho$}] at (7.5, 4.15) {};
	\node[label=right:{$\binom{5}{2}$ \; $5 L \rho$}] at (7.5, 3.45) {};
	\node[label=right:{$\binom{4}{2}$ \; $4 L \rho$}] at (7.5, 2.75) {};
	\node[label=right:{$\binom{3}{2}$ \; $3 L \rho$}] at (7.5, 1.95) {};
	\node[label=right:{$\binom{2}{2}$ \; $2 L \rho$}] at (7.5, 1.35) {};
	\node[label=right:{$\binom{3}{2}$ \; $3 L \rho$}] at (7.5, 0.6) {};

	% Gray dashed lines to visually separate holding times
	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2);
	\draw[color=gray, dashed] (7.5, 1.5) -- (11, 1.5);
	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4);
	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1);
	\draw[color=gray, dashed] (7.5, 3.8) -- (11, 3.8);
	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5);
	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4);
	\draw[color=gray, dashed] (7.5, 6) -- (11, 6);
	\draw[color=gray, dashed] (7.5, 6.6) -- (11, 6.6);
	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2);
\end{tikzpicture}
\end{tabular}
}
\caption{(A)
A realisation of the graph traversed by Hudson's algorithm started from a
sample of three continuous chromosomes of length $L$ at time $t = 0$, and
propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
at event B, while that on $[0, v)$ is reached at event C.
The non-ancestral segment $[v, w)$ above
A contributes to the rate of effective recombinations because it
is trapped between ancestral segments. The two columns titled CA and RE
are the respective rates of mergers and recombinations when
the recombination rate is $\rho$.
(B) A corresponding realisation of a big ARG, which augments Hudson's algorithm
by tracking nonancestral lineages. The result is a simpler state space and
dynamics, at the cost of extra nodes and edges, highlighted in red, which do
not affect the local tree at any site.}
\label{hudson_vs_bigARG}
\end{figure}

@@

In the Griffiths formulation, each edge in the graph corresponds to an extant
lineage and nodes are events in the process (the initial $n$ leaf nodes are
``sampling'' events). As in Hudson's
algorithm, common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events are different
to Hudson's algorithm, however. Rather than occuring at a rate which depends on
the specific distribution of ancestral material among lineages, the
rate of recombination is now simply $k \rho (m - 1)$. At a recombination event
we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < L$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMRCA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. The graph structure and breakpoints associated with
recombination nodes provides sufficient information to later recover the marginal
trees (see the next section for more details and discussion on this point).

@@

\section*{The Event ARG data structure}
% When did we move from the ARG is a process to the ARG is a data structure?
% What was the historical progression from one to the other?
Early work on ARG inference focused on the problem of
of inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
% FIXME: need some proper statistical language here to describe why these
% methods didn't work so well.
These methods met with limited success
due to the overwhelmingly large and
horribly structured [Jere: what's the right way to say this?] state space of ARGs.
The focus subsequently shifted to the more
tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset [CITATIONS], and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
The distinction between the ARG as a stochastic process
and the ARG as a \emph{data structure}~\citep{minichiello2006mapping}
became important at this point,
since parsimonious ARGs have no immediate connection to
the stochastic process. Indeed, the ARG instances produced by
the coalescent with recombination are usually highly unparsimonious,
since they contain many invisible recombinations
(although those produced under the SMC may be parsimonious, depending
on the pattern of mutations that is present).

@@

% it's not reasonable to assume that one type of event happens
% at a time any more.
The Griffiths ARG encoding is also limited in the types of ancestral
history that can be described. This ARG encoding is derived from the coalescent
with recombination, which explicitly assumes that only one type
of event can occur (i.e., common ancestry or recombination) at a
time. This model is derived as an approximation to a Wright-Fisher
model, assuming that the sample size $n$ is much less than the
population size $N$ and [some assumptions about genome length]
\textcolor{red}{what is this Jere?}. While these approximations are
reasonable
to make when studying the Kreitman \emph{Drosphila Melanogaster}
dataset~\citep{kreitman1983nucleotide} ($n=11$, $N\approx 10^6$,
at 43 variant sites over x Kb \textcolor{red}{details and citation
for Ne estimate}) that is classically used for benchmarking
ARG inference methods [cites],
they are highly questionable in modern datasets. % absurd really
For example, there are now many
human datasets consisting of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and thus our sample size is often much \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
% TODO make this better. Ask Gregor for some refs?
% Basically, Ag datasets make the assumption of isolated
% rare recombinations absurd.
Agricultural datasets such as the 1000 bull genomes~\citep{hayes20191000}
combine genome sequence data with detailed pedigrees, with
sample data from multiple generations.
Complete chromosome-level assemblies are now possible
in humans~\citep{miga2020telomere},
and projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
While the coalescent model can be suprisingly robust to such
basic violations to the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.
In a Wright-Fisher model, coalescence and recombination can
occur at the same time in the same genomes, and we can be
certain that this is also true in the incredibly rich
datasets that are available today.

>>>>

In this paper we [turn into proper round up paragraph later]
\begin{enumerate}
\item Discuss the classical Ancestral Recombination Graph formulation,
and show how it derives from (and is limited by) the coalescent with
recombination.
\item Suggest a definition of ARGs that is free of these limitations,
where the formulation is derived from a (normally diploid) pedigree rather than from
coalescent approximations, and where the genetic
material which is transmitted is described by annotations attached to
edges rather than nodes.
\item Suggest a classification of ARG nodes that helps us to understand
the properties of the structures inferred by different methods, and illustrate
with some examples.
% Perhaps we should also mention something about simplification and unknowable
% nodes, e.g. "show how some of the nodes are not required for useful interpretation
% of ancestry".
\item Throughout, we consider ARGs defined for a single chromosome comprising
of a fixed number of discrete sites. Our argument applies equally well to ARGs
defined for several non-homologous chromosomes \citep{fearnhead2003ancestral,
koskela2019robust} and to idealised, continuous chromosomes, but addressing 
all these cases simultaneously would result in needlessly bulky notation.  
\end{enumerate}

@@

	\draw (0, 0) -- (0, 1.7) -- (-1,1.7) -- (1,1.7);
	\draw (-1.4, 1.5) -- (-1.1, 1.5) -- (-1.1, 1.4) -- (-1.4, 1.4) -- (-1.4, 1.5);
	\draw (-1.1, 1.45) -- (-0.6, 1.45);
	\draw (0.6, 1.45) -- (0.9, 1.45);
	\draw (0.9, 1.5) -- (1.4, 1.5) -- (1.4, 1.4) -- (0.9, 1.4) -- (0.9, 1.5);
	\draw (1.2, 1.6) -- (1.2, 1.3);
	\node [label=below:{\small $x$}] at (1.2, 1.5) {};

@@

	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
	\draw (-0.4, 2.85) -- (0.2, 2.85);
	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
	\draw (1.6, 2.85) -- (1.9, 2.85);
	\draw (1.9, 2.9) -- (2.2, 2.9) -- (2.2, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
	\draw (2.2, 2.85) -- (2.4, 2.85);

	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
	\node [scale=0.2,label=above left:{\small A}] at (-0.5,4.5) {A};
	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
	\draw (-0.6, 4.25) -- (-0.3, 4.25);
	\draw (-0.3, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.3, 4.2) -- (-0.3, 4.3);
	\draw (-0.5, 4.4) -- (-0.5, 4.1);
	\node [label=below:{\small $y$}] at (-0.5, 4.3) {};

@@

	% Edge annotations above each event
	\node [label=left:{\small$\{(0,m,1)\}$}] at (0.2,0.6) {};
	\node [label=left:{\small$\{(0,m,1)\}$}] at (4.2,0.4) {};
	\node [label=right:{\small$\{(0,m,1)\}$}] at (5.8,0.4) {};
	\node [label=left:{\small$\{(0,m,2)\}$}] at (5.2,1.6) {};
	\node [label=left:{\small$\{(0,v,1)\}$}] at (-0.8,2.3) {};
	\node [label=right:{\small$\{(v,m,1)\}$}] at (0.8,2.1) {};
	\node [label=right:{\small$\{(0,w,2)\}$}] at (3.8,3.9) {};
	\node [label=left:{\small$\{(w,m,2)\}$}] at (6.2,5) {};
	\node [label=right:{\small$\{(x,m,1)\}$}] at (-0.2,3.8) {};
	\node [label=right:{\small$\{(v,x,1)\}$}] at (1.8,4.1) {};
	\node [label=left:{\small$\{(0,v,1), (x,m,1)\}$}] at (-0.3,5.2) {};
	\node [label=right:{\small$\{(0,v,2)\}$}] at (2.8,6.4) {};
	\node [label=left:{\small$\{(x,m,1)\}$}] at (-1.3,7) {};
	\node [label=right:{\small$\{(0,v,1)\}$}] at (0.3,6.4) {};

@@

	% Numbers of extant ancestors and links, from top to bottom
	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
	\node[label=right:{$\binom{2}{2}$ \; $\frac{2 m - x - w - 2}{m - 1} \rho$}] at (7.5, 7.6) {};
	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + 2 v - x - w - 3}{m - 1} \rho$}] at (7.5, 6.6) {};
	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m + v - w - 3}{m - 1} \rho$}] at (7.5, 5.7) {};
	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m + x - v - 4}{m - 1} \rho$}] at (7.5, 4.95) {};
	\node[label=right:{$\binom{5}{2}$ \; $\frac{2 m - 5}{m - 1} \rho$}] at (7.5, 3.8) {};
	\node[label=right:{$\binom{4}{2}$ \; $\frac{2 m - 4}{m -1} \rho$}] at (7.5, 2.75) {};
	\node[label=right:{$\binom{3}{2}$ \; $\frac{2 m - 3}{m - 1} \rho$}] at (7.5, 2.05) {};
	\node[label=right:{$\binom{2}{2}$ \; $2 \rho$}] at (7.5, 1.45) {};
	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 0.6) {};

	% Gray dashed lines to visually separate holding times
	\draw[color=gray, dashed] (7.5, 1.2) -- (11.1, 1.2);
	\draw[color=gray, dashed] (7.5, 1.7) -- (11.1, 1.7);
	\draw[color=gray, dashed] (7.5, 2.4) -- (11.1, 2.4);
	\draw[color=gray, dashed] (7.5, 3.1) -- (11.1, 3.1);
	\draw[color=gray, dashed] (7.5, 4.5) -- (11.1, 4.5);
	\draw[color=gray, dashed] (7.5, 5.4) -- (11.1, 5.4);
	\draw[color=gray, dashed] (7.5, 6) -- (11.1, 6);
	\draw[color=gray, dashed] (7.5, 7.2) -- (11.1, 7.2);
\end{tikzpicture}
&
\begin{tikzpicture}
	\draw (-0.4, -0.2) -- (0.4, -0.2) -- (0.4, -0.3) -- (-0.4, -0.3) -- (-0.4, -0.2);
	\draw (-0.1, -0.1) -- (-0.1, -0.4);
	\node [label=below:{\small $v$}] at (-0.1, -0.2) {};
	\draw (3.6, -0.2) -- (4.4, -0.2) -- (4.4, -0.3) -- (3.6, -0.3) -- (3.6, -0.2);
	\draw (5.6, -0.2) -- (6.4, -0.2) -- (6.4, -0.3) -- (5.6, -0.3) -- (5.6, -0.2);

@@

	\draw (1, 1.7) -- (1, 3.1) -- (0,3.1) -- (2,3.1);
	\draw (-0.4, 2.85) -- (0.2, 2.85);
	\draw (0.2, 2.9) -- (0.4, 2.9) -- (0.4, 2.8) -- (0.2, 2.8) -- (0.2, 2.9);
	\draw (1.6, 2.85) -- (1.9, 2.85);
	\draw (1.9, 2.9) -- (2.2, 2.9) -- (2.2, 2.8) -- (1.9, 2.8) -- (1.9, 2.9);
	\draw (2.2, 2.85) -- (2.4, 2.85);

	\draw (-1, 1.7) -- (-1, 4.5) -- (0,4.5) -- (0,3.1);
	\draw (-0.9, 4.3) -- (-0.6, 4.3) -- (-0.6, 4.2) -- (-0.9, 4.2) -- (-0.9, 4.3);
	\draw (-0.6, 4.25) -- (-0.3, 4.25);
	\draw (-0.3, 4.3) -- (-0.1, 4.3) -- (-0.1, 4.2) -- (-0.3, 4.2) -- (-0.3, 4.3);
	\draw (-0.5, 4.4) -- (-0.5, 4.1);
	\node [label=below:{\small $y$}] at (-0.5, 4.3) {};

@@

	% Numbers of extant ancestors and links, from top to bottom
	\node[label=right:{CA \; RE}] at (7.5, 8.2) {};
	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 7.6) {};
	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 6.9) {};
	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 6.3) {};
	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 5.7) {};
	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 4.95) {};
	\node[label=right:{$\binom{6}{2}$ \; $6 \rho$}] at (7.5, 4.15) {};
	\node[label=right:{$\binom{5}{2}$ \; $5 \rho$}] at (7.5, 3.45) {};
	\node[label=right:{$\binom{4}{2}$ \; $4 \rho$}] at (7.5, 2.75) {};
	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 2.05) {};
	\node[label=right:{$\binom{2}{2}$ \; $2 \rho$}] at (7.5, 1.45) {};
	\node[label=right:{$\binom{3}{2}$ \; $3 \rho$}] at (7.5, 0.6) {};

	% Gray dashed lines to visually separate holding times
	\draw[color=gray, dashed] (7.5, 1.2) -- (11, 1.2);
	\draw[color=gray, dashed] (7.5, 1.7) -- (11, 1.7);
	\draw[color=gray, dashed] (7.5, 2.4) -- (11, 2.4);
	\draw[color=gray, dashed] (7.5, 3.1) -- (11, 3.1);
	\draw[color=gray, dashed] (7.5, 3.8) -- (11, 3.8);
	\draw[color=gray, dashed] (7.5, 4.5) -- (11, 4.5);
	\draw[color=gray, dashed] (7.5, 5.4) -- (11, 5.4);
	\draw[color=gray, dashed] (7.5, 6) -- (11, 6);
	\draw[color=gray, dashed] (7.5, 6.6) -- (11, 6.6);
	\draw[color=gray, dashed] (7.5, 7.2) -- (11, 7.2);
\end{tikzpicture}
\end{tabular}
}
\caption{(A)
A realisation of the graph traversed by Hudson's algorithm started from a
sample of three chromosomes of length $m$ at time $t = 0$, and
propagated until time $T$. The MRCA on the genetic interval $[v, w)$ is reached
at event B, while that on $[0, v)$ is reached at event C.
The non-ancestral segment $[v, w)$ above
A contributes to the rate of effective recombinations because it
is trapped between ancestral segments. The two columns titled CA and RE
are the respective rates of mergers and recombinations when
the recombination rate is $\rho$.
(B) A corresponding realisation of a big ARG, which augments Hudson's algorithm
by tracking nonancestral lineages. The result is a simpler state space and
dynamics, at the cost of extra nodes and edges, highlighted in red, which do
not affect the local tree at any site.}
\label{hudson_vs_bigARG}
\end{figure}

@@

In the Griffiths formulation, each edge in the graph corresponds to an extant
lineage and nodes are events in the process (the initial $n$ leaf nodes are
``sampling'' events). As in Hudson's
algorithm, common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events are different
to Hudson's algorithm, however. Rather than occuring at a rate which depends on
the specific distribution of ancestral material among lineages, the
rate of recombination is now simply $k \rho$. At a recombination event
we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMRCA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. The graph structure and breakpoints associated with
recombination nodes provides sufficient information to later recover the marginal
trees (see the next section for more details and discussion on this point).

@@

\section*{The Event ARG data structure}
% When did we move from the ARG is a process to the ARG is a data structure?
% What was the historical progression from one to the other?
Early work on ARG inference focused on the problem of
of inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
% FIXME: need some proper statistical language here to describe why these
% methods didn't work so well.
These methods met with limited success
due to the overwhelmingly large state space of ARGs, which also has
no simple geometry or neighbourhood structure that an inference or
sampling method could exploit.
The focus subsequently shifted to the more
tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset [CITATIONS], and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
The distinction between the ARG as a stochastic process
and the ARG as a \emph{data structure}~\citep{minichiello2006mapping}
became important at this point,
since parsimonious ARGs have no immediate connection to
the stochastic process. Indeed, the ARG instances produced by
the coalescent with recombination are usually highly unparsimonious,
since they contain many invisible recombinations
(although those produced under the SMC may be parsimonious, depending
on the pattern of mutations that is present).

@@

% it's not reasonable to assume that one type of event happens
% at a time any more.
The Griffiths ARG encoding is also limited in the types of ancestral
history that can be described. This ARG encoding is derived from the coalescent
with recombination, which explicitly assumes that only one type
of event can occur (i.e., common ancestry or recombination) at a
time. This model is derived as an approximation to a Wright-Fisher
model, assuming that the sample size $n$ is much less than the
population size $N$, and that the genome is short enough that,
with high probability, the number of concurrent lineages in the ARG remains
much smaller than $N$ at each time. While these approximations are
reasonable
to make when studying the Kreitman \emph{Drosphila Melanogaster}
dataset~\citep{kreitman1983nucleotide} ($n=11$, $N\approx 10^6$,
at 43 variant sites over x Kb \textcolor{red}{details and citation
for Ne estimate}) that is classically used for benchmarking
ARG inference methods [cites],
they are highly questionable in modern datasets. % absurd really
For example, there are now many
human datasets consisting of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and thus our sample size is often much \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
% TODO make this better. Ask Gregor for some refs?
% Basically, Ag datasets make the assumption of isolated
% rare recombinations absurd.
Agricultural datasets such as the 1000 bull genomes~\citep{hayes20191000}
combine genome sequence data with detailed pedigrees, with
sample data from multiple generations.
Complete chromosome-level assemblies are now possible
in humans~\citep{miga2020telomere},
and projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
While the coalescent model can be suprisingly robust to such
basic violations to the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.
In a Wright-Fisher model, coalescence and recombination can
occur at the same time in the same genomes, and we can be
certain that this is also true in the incredibly rich
datasets that are available today.
